A Mountaintop View Requires Minimal Sorting: \break A Faster Contour Tree Algorithm

### 2014-10-31 paint-contour-tree.tex
section 8 modifications

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\fcol}{full}
\newcommand{\cur}{cur}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\h}{att}
\newcommand{\init}{{\tt init}}
\newcommand{\jc}{\cJ_C}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\merge}{{\tt merge}}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pal}{P}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{\cC}
\newcommand{\redH}{\widetilde{H}}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

We present the first algorithm that can avoid the global sort and has a refined time complexity that
depends on the contour tree structure. Intuitively, if the tree is short and fat, we get significant improvements 
in running time. For a partition of the contour tree into a set of descending paths, $P$,
our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t))$. Moreover, we prove a general lower bound of
$\Omega(\sum_{p \in P} |p|\log |p|)$ for computing contour trees, showing
that the tree structure must play a role in the time complexity.

@@

Massive geometric data is often represented as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some mesh (simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a specific topological data structure used to represent and visualize
the function $f$.

It is convenient to think of $f$ as a $(d+1)$-dimensional manifold, or alternately consider
$\MM$ in $\RR^{d+1}$ with the last coordinate (height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. At the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other,
split into new components, contours acquire or lose handles. These are the relevant events where 
the topology of the level set changes. The contour tree is a concise representation of all these events.

@@

\Fig{relative} and \Fig{sorting} show examples of simplicial complices with heights and their contour trees.
Think of the edges pointing downwards, so we can talk of the ``up-degree" and ``down-degree" of vertices.
Leaves of down-degree $1$ are maxima, because a new contour is created at such point. Leaves of up-degree
$1$ are minimum, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
merge, and down-degree $2$ vertices are ``splits". In these examples ($d=2$), contours are closed loops.
In dimensions $d \geq 3$, we may have internal
vertices of both up and down-degree $1$. They correspond to contours (which are $2$-manifolds) gaining or losing handles.

@@

But this is only a worst-case bound. All previous algorithm begin by sorting (at least) the critical points, and
pay must $\Omega(t\log t)$ for all instances. Can we beat this sorting bound? Is there a more refined version of the complexity
of computing contour trees? Our main result gives an affirmative answer. First, we require some technical definitions.

In a contour tree $T$, a \emph{leaf path} is simply a contiguous path in $T$
containing a leaf, which is also monotonic in the height values of its vertices. A leaf path runs either upwards
or downwards, never both.
% Then in the following two subsections 
%we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.

@@

\begin{theorem} \label{thm:main-alg} Consider PL-Morse\footnote{This is a standard ``well-behaved" assumption that implies
that all critical points have distinct values, and all the contour tree has max-degree $3$. Our algorithm
does not need this, but it simplifies the proof and presentation.} $f:\MM \mapsto \RR$, where the contour tree $T$
has maximum degree $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implictly by the algorithm).
\end{theorem}

Let us interpret the running time. 
For any path decomposition $P(T)$ and any $p \in P(T)$, obviously $|p| \leq |T| = t$.
So the running time is clearly $O(t\log t + N)$. But this is only the case when $T$ is tall and thin.  
An obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $P(T)$. 
This is clearly an improvement when the tree is short and fat.
But if $P(T)$ is a rooted, balanced, binary tree, then $\sum_{p \in P(T)} |p|\log |p| = O(t)$.
Our running time is now $O(t\alpha(t))$, a significant improvement from $O(t\log t)$.
Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
(In the example, $t = \Omega(n)$.) The running time promised by \Thm{main-alg}
is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$.

We complement \Thm{main-alg} with an almost matching lower bound. We show that the complexity
of $\sum_{p \in P(T)} |p|\log |p|$ is required for any contour tree algorithm. The exact phrasing
of the lower bound requires some care. Consider the $d=2$ (terrain) case. For any $f$,
we use $\cost(f)$ to denote $\sum_{p \in P(T)} |p|\log |p|$ for the specific $P(T)$
given in \Thm{main-alg}. Ideally, we would like to argue that any contour-tree algorithm
must pay $\cost(f)$ for $f$. But this is obviously false, since one could consider an algorithm
that hard codes the answer for $f$ and merely checks this answer.

Instead, call a possible cost $C$ \emph{valid} if $C = \cost(f)$ for some $f$\footnote{If one is concerned
that $C$ might be irrational, define valid if $2^C = 2^{\cost(f)}$.}. We prove
a refined lower bound matching \Thm{main-alg}. A \emph{height decision tree} is an algorithm
that is allowed aribtrary computation on the domain $\MM$, and is an algebraic decision
tree over the heights. The running time is simply the number of queries made.

\begin{theorem} \label{thm:main-lb} Consider the set of functions $\bF_C$ that all have the same (valid) cost $C$.
Any height decision tree that correctly computes
the contour tree on all of $\bF$ has an average case (uniform over $\bF$) running time of $\Omega(C)$.
\end{theorem}

In other words, it is not possible to improve on \Thm{main-alg} even when restricted to the subset
of inputs of fixed cost.

@@

There has been a long series of improved algorithms for computing contour trees.
van Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case
The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N)$.
One cannot overstate the importance of this result, since it forms the basis 
of subsequent practical and theoretical work in computational topology applications. 
Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an unconditional $O(N+t\log t)$ algorithm.

@@

\begin{figure}[h]\centering
    \includegraphics[width=.7\linewidth,height=.3\linewidth]{figs/balanced3}
    \vspace{.1in}

@@

This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
has a well-defined height, and this includes the infinite facet. 
We think of $d$ as constant, and assume that $\MM$ is represented in data structures that allow constant-time traversals
in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a DCEL for higher dimensions.)
Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$.
We can think of $f(x)$ as the ``height" of a point $x \in \MM$, and encoded in the representation of $\MM$. 
%Abusing notation, $\MM$ refers to this encoding, so we do not refer to $f$.

@@

\begin{definition} \label{def:tree} The contour tree is the graph on vertex set $\cV$, where
edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
there is an edge $(v_i,v_j)$. (Conventionally, edges are directed from higher function value to lower.)
\end{definition}

We denote the contour tree of $\MM$
by $\reeb(\MM)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
It is not immediately obvious that this graph is a tree, but the above definition is equivalent to that of the Reeb graph of $f:\MM \rightarrow \RR$, 
known to be a tree for PL functions defined over PL manifolds (Chapter 6 of~\cite{HaEd10}). Alternate definitions of the contour tree
in~\cite{csa-cctad-00} imply this is a tree. We can talk about up-degrees and down-degrees in $\reeb(\MM)$
We assume there are no ``multi-saddles", so up and down-degrees are at most $2$, and the total degree is at most $3$. 
This is again a standard assumption in topological algorithms and can be achieved by vertex unfolding (Section 6.3 in~\cite{HaEd10}).

@@

The $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$ given by \Def{deg} might actually be the same
contour in $f$. The up-degree is defined as the number of $(f(x)+\delta)$-contours
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, then the level set gains a handle.
When $d=2$, this distinction is not necessary, since any point with $>1$
Morse degrees will have some $>1$ degree in $\reeb(\MM)$.

@@

\smallskip
\begin{asparaenum}
	\item Let $\MM' = \cut(\phi,\MM)$.
	\item Construct $\reeb(\MM')$ and let $A, B$ be the nodes corresponding to the new boundaries created
	in $\MM'$. (One is a minima and the other is maxima.)
	\item Since $A, B$ are leaves, they each have unique neighbors $A'$ and $B'$, respectively. Insert
	edge $(A',B')$ and delete $A, B$ to obtain $\reeb(\MM)$.
\end{asparaenum}
\end{minipage}}

@@

\begin{lemma} \label{lem:cut} Consider a regular contour $\phi$ contained in a contour class (of an edge of $\reeb(\MM))$
$(u,v)$ and denote $\MM' = \cut(\phi,\MM)$. Then $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup \cV(\MM)$
and $\cE(\reeb(\MM')) = \{(u,\phi^+),(\phi^-,v)\} \cup (\cE(\MM) \setminus (u,v))$.
\end{lemma}

@@

\begin{theorem} \label{thm:jordan} $\cut(\phi,\MM)$ consists of two
disconnected simplicial complexes.
\end{theorem}

\begin{proof} Denote (as in \Lem{cut}) the edge containing $\phi$ to be $(u,v)$. Suppose for contradiction that there is a path between vertices $u$ and $v$
in $\MM' = \cut(\phi,\MM)$. By \Thm{carr-path}, there is a path in $\cC(\MM')$ between $u$ and $v$. Since $\phi^+$ and $\phi^-$
are leaves in $\cC(\MM')$, this path obviously cannot use edges incident to them. By \Lem{cut},
all the edges of this path are in $\cE(\cC(\MM)) \setminus (u,v)$. So we get a cycle in $\cC(\MM)$, a contradiction.
To show that there are exactly two connected components in $\cut(\phi,\MM)$, it suffices
to see that $\cC(\MM')$ has two connected components (by \Lem{cut}) and apply \Thm{carr-path}.
\end{proof}

@@

\begin{figure}[h]\centering
    \includegraphics[width=.18\linewidth,height=.35\linewidth]{figs/orderGray1}%
    \hspace{.8in}
    \includegraphics[width=.18\linewidth,height=.35\linewidth]{figs/orderGray2}
    \caption{On left, downward paint spilling only (each shade of gray represents a different color), producing a grid.  On right, flipping the direction of paint spilling.}
    \label{fig:order}
\end{figure}

@@

\smallskip
\begin{compactenum}
	\item Determine interface of $\wet(x,\NN)$. 
	\item If the interface is empty, simply output $\NN$. Otherwise, denote the contours by $\phi_1, \phi_2, \ldots, \phi_k$ and set $\phi'_i = \lift(\phi_i)$.
	\item Initialize $\NN_1 = \NN$.
	\item For $i$ from $1$ to $k$: 
	\begin{compactenum}
		\item Construct $\cut(\phi'_i,\NN_i)$, consisting of dry complex $\LL_i$ and remainder $\NN_{i+1}$.
		\item Let the newly created boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively
		call $\rain(B_i,\LL_i)$.
	\end{compactenum}
	\item Output $\NN_{k+1}$ together with any complexes output by recursive calls.
\end{compactenum}
\end{minipage}}

@@

Consider the first time that $F$ is cut, say, during the call to $\rain(x,\NN)$.
Specifically, say this happens when $\cut(\phi'_i,\NN_i)$ is constructed. 
So $F$ is cut by some hyperplane into lower and upper portions (which are then triangulated).
Since the cut is infinitesimally above an interface, the lower portion is basically wet.
It can never be part of $\LL_1, \LL_2, \ldots$, which go into recursive calls. This portion
is in $\NN_{k+1}$. The upper portion/facet (call it $U$) is in $\LL_i$. Note that the lower
boundary of $U$ is in the boundary $B_i$. Since a recursive call is made to $\rain(B_i,\LL_i)$
(and $\LL_i$ is inverted), $U$ becomes wet. Hence, $U$ will not be cut subsequently.
All in all, $F$ is cut at most once.
\ignore{
As the number of edges is $O(n)$, arguing that there is at most one boundary vertex created along any given edge will imply the claim.
Consider some edge $e$ and take the first time a boundary vertex is created in $e$.
This happens when some contour $\phi$ intersects $e$. On applying $\cut(\NN,\lift(\phi))$,
consider the dry manifold $\LL$ and wet manifold $\MM$ constructed. The edge $e$ is split into two parts, the dry and wet part.
The wet part is in $\MM$ and will never be involved in any recursive call.

@@

\section{Painting to compute contour trees}

@@

\begin{definition}
 A \emph{leaf assignment} $\chi$ of a tree $T$ assigns \emph{two} distinct leaves to each internal vertex $v$,
 one from the left child and one from the right child subtree of $v$ (naturally if $v$ has only one child it is assigned 
 only one color).
\end{definition}

@@

\newcommand{\pathTree}{P_{\mathcal{S}}}
\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}
\section{Leaf assignments and path decompositions}

@@

\subsection{Join Trees}
In this section we show that any algorithm for computing the join tree $T$ must pay at least $f_{path}(P(T))$ (see \Def{path}) for any path decomposition $P(T)$.

%We say two shrubs $S_1$ and $S_2$ (of join trees of potentially different surfaces) are equivalent if there is an isomorphism, $f$,
%between the two shrubs such that for any node $v\in S_1$ corresponding to a path of size $|v|$, $f(v)$ also corresponds to a path of size $|v|$.
%For the purposes of the next lemma, we say two $2$-manifolds are different, if their corresponding contour trees are the different.

In the following note that two distinct PL surfaces can have the same graph structure defining their meshes, but different height values associated with their vertices.

@@

\begin{lemma}
\label{lem:pathDecomp}
Let $T$ be a join tree, and let $P = P(T)$ any path decomposition of $T$.  %, and $\pathTree$ the corresponding shrub.  
Then there is a set of at least $\Pi_{p_i\in P} (|p_i|-1)!$ distinct PL $2$-manifolds, all of which have the same defining mesh.
\end{lemma}
\begin{proof}
 We describe a simple mesh, with heights specified for all vertices except for the saddles.  The main idea is that 
 this mesh will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of join trees 
 whose size is that described in the lemma statement.

>>>>

\newcommand{\build}{{\tt build}}
\newcommand{\col}{col}
\newcommand{\fcol}{full}
\newcommand{\cur}{cur}
\newcommand{\cut}{{\tt cut}}
\newcommand{\cont}{\psi}
\newcommand{\cost}{\mathop{cost}}
\newcommand{\costs}{\mathop{costs}}
\newcommand{\h}{att}
\newcommand{\init}{{\tt init}}
\newcommand{\jc}{\cJ_C}
\newcommand{\lift}{{\tt lift}}
\newcommand{\mcol}{mcol}
\newcommand{\merge}{{\tt merge}}
\newcommand{\onestep}{{\tt piece}}
\newcommand{\paint}{{\tt paint}}
\newcommand{\pal}{P}
\newcommand{\pmax}{P_{\max}}
\newcommand{\rain}{{\tt rain}}
\newcommand{\reeb}{\cC}
\newcommand{\redH}{\widetilde{H}}
\newcommand{\rep}{rep}
\newcommand{\st}{K}
\newcommand{\surgery}{{\tt surgery}}
\newcommand{\touch}{T}
\newcommand{\update}{{\tt update}}
\newcommand{\wet}{{\tt wet}}

@@

We present the first algorithm that can avoid the global sort and has a refined time complexity that
depends on the contour tree structure. Intuitively, if the tree is short and fat, we get significant improvements 
in running time. For a partition of the contour tree into a set of descending paths, $P$,
our algorithm runs in $O(\sum_{p\in P} |p|\log|p| + t\alpha(t)+N)$, which is 
always bounded by $O(t\log t + N)$ and is $O(N+t\alpha(t))$ for balanced trees. 
Moreover, we prove a general lower bound of
$\Omega(\sum_{p \in P} |p|\log |p|)$ for computing contour trees, showing
that the tree structure must play a role in the time complexity.

@@

Massive geometric data is often represented as a function $f: \RR^d \mapsto \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some mesh (i.e. simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a specific topological data structure used to represent and visualize
the function $f$.

It is convenient to think of $f$ as a $(d+1)$-dimensional manifold, or alternately consider
$\MM$ in $\RR^{d+1}$ with the last coordinate (i.e. height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other or
split into new components, contours acquire or lose handles. These are the relevant events where 
the topology of the level set changes. The contour tree is a concise representation of all these events.

@@

\Fig{relative} and \Fig{sorting} show examples of simplicial complexes, with heights and their contour trees.
Think of the contour tree edges as pointing downwards, so we can talk of the ``up-degree" and ``down-degree" of vertices.
Leaves of down-degree $1$ are maxima, because a new contour is created at such points. Leaves of up-degree
$1$ are minima, because a contour is destroyed there. Up-degree $2$ vertices are ``joins", where contours
merge, and down-degree $2$ vertices are ``splits".  When $d=2$ (as in \Fig{relative}), contours are closed loops.
In dimensions $d \geq 3$, we may have internal
vertices of both up and down-degree $1$. They correspond to contours (which are $2$-manifolds) gaining or losing handles.

@@

But this is only a worst-case bound. All previous algorithm begin by sorting (at least) the critical points, and
hence must pay $\Omega(t\log t)$ for all instances. Can we beat this sorting bound? Is there a more refined version of the complexity
of computing contour trees? Our main result gives an affirmative answer. First, we require some technical definitions.

In a contour tree $T$, a \emph{leaf path} is simply a contiguous path in $T$
containing a leaf, which is also monotonic in the height values of its vertices. 
%A leaf path runs either upwards or downwards, never both.
%
% Then in the following two subsections 
%we relate path decomposition costs (defined below) to the running time of our algorithm using $\Lem{algCost}$.

@@

\begin{theorem} \label{thm:main-alg} Consider a PL-Morse\footnote{This is a standard ``well-behaved" assumption that implies
that all critical points have distinct values, and the contour tree has max-degree $3$. Our algorithm
does not need this, but it simplifies the proof and presentation.} $f:\MM \mapsto \RR$, where the contour tree $T$
has maximum degree $3$. There is an algorithm whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
\end{theorem}

Let us interpret the running time. 
For any path decomposition $P(T)$, obviously $\sum_{p\in P(T)}|p| = |T| = t$, and 
so the running time is $O(t\log t + N)$. But this is tight only when $T$ is tall and thin.  
A better but still obvious upper bound is $O(t\log D + t\alpha(t) + N)$, where $D$ is the diameter of $P(T)$. 
This is clearly an improvement when the tree is short and fat.
In particular, for any decomposition of a rooted balanced binary tree, this gives $O(t\log\log n + t\alpha(t) + N)$, a significant improvement from $O(t\log t+N)$.
However, our actual bound in this case gives an even further improved $O(t\alpha(t)+N)$.
%
Consider the example of \Fig{sorting} extended to a larger binary tree both above and below.
(In the example, $t = \Omega(n)$.) The running time promised by \Thm{main-alg}
is $O(n\alpha(n))$, whereas all previous algorithms require $\Omega(n\log n)$.

We complement \Thm{main-alg} with an almost matching lower bound, showing that the complexity
of $\sum_{p \in P(T)} |p|\log |p|$ is required for any contour tree algorithm. The exact phrasing
of the lower bound requires some care. Consider the $d=2$ (terrain) case, and fix a mesh $\MM$.
As referenced in \Thm{main-alg}, for any height function $f$ our algorithm defines a 
specific path decomposition $P(T)$\footnote{The specific decomposition used is defined in \Sec{pathDecomp}, 
but also depends on the raining and painting procedures described in \Sec{rain} and \Sec{paint}.  
The important thing is that our algorithm is deterministic, and so for a fixed $\MM$ and $f$ it is well defined.}. 
We define $\cost(f) = \sum_{p \in P(T)} |p|\log |p|$ for this specific decomposition $P(T)$.
Ideally, we would like to argue that any contour-tree algorithm
must pay $\cost(f)$ for $f$. 
But this is obviously false, since one could consider an algorithm
that hard codes the answer for $f$ and merely checks this answer.

Instead, call a possible cost $C$ \emph{valid} if $C = \cost(f)$ for some $f$ 
defined over the fixed mesh $\MM$\footnote{If one is concerned
that $C$ might be irrational, define valid if $2^C = 2^{\cost(f)}$.}. We prove
a refined lower bound matching \Thm{main-alg}.

\begin{definition}\label{def:hfdt}
A \emph{height function decision tree} is an algorithm
that is allowed arbitrary computation on the domain $\MM$, and is an algebraic decision
tree over the heights. The running time is simply the number of queries made.
\end{definition}

\begin{theorem} \label{thm:main-lb} Consider the set of functions $\bF_C$ that all have the same (valid) cost $C$ 
(defined over a fixed mesh $\MM$).  Any height function decision tree that correctly computes
the contour tree on all of $\bF_C$ has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
\end{theorem}

% We complement \Thm{main-alg} with an almost matching lower bound, showing that the complexity
% of $\sum_{p \in P(T)} |p|\log |p|$ is required for any contour tree algorithm. The exact phrasing
% of the lower bound requires some care. Consider the $d=2$ (terrain) case, and fix a mesh $\MM$.
% Now any height function $f$ (together with $\MM$) defines a specific contour tree $T$.
% As referenced in \Thm{main-alg}, given $f$, our algorithm then defines some specific path 
% decomposition $\overline{P}(T)$, such that the running time of our algorithm is bounded by 
% $O(\cost(\overline{P}(T))+t\alpha(t)+N)$, where $\cost(P(T)) = \sum_{p \in P(T)} |p|\log |p|$.
% A priori, which path decomposition this will be is unclear\footnote{The specific decomposition, defined in \Sec{pathDecomp}, 
% depends on arbitrary choices made during the raining and painting procedures described in \Sec{rain} and \Sec{paint}.}, 
% however we know there is some large set of potential path decompositions it lies in
% (the exact set is defined in \Def{path2}, but for now one can think of the set of all decompositions).
% Let $\costs(f)$ denote the set of costs of all these potential path decompositions that $f$ defines
% (i.e. $\cost(\overline{P}(T))\in \costs(f)$).
% 
% Ideally, we would like to argue that any contour-tree algorithm must pay $\cost(\overline{P}(T))$ for $f$. 
% But this is obviously false, since one could consider an algorithm
% that hard codes the answer for $f$ and merely checks this answer.
% Instead, call a possible cost $C$ \emph{valid} if $C \in \costs(f)$ for some $f$ 
% defined over the fixed mesh $\MM$\footnote{If one is concerned
% that $C$ might be irrational, define valid if $2^C = 2^{\cost(f)}$.}. We prove
% a refined lower bound matching \Thm{main-alg}. 
% 
% \begin{definition} \label{def:hfdt}
% A \emph{height function decision tree} is an algorithm
% that is allowed arbitrary computation on the domain $\MM$, and is an algebraic decision
% tree over the heights. The running time is simply the number of queries made.
% \end{definition}
% 
% \begin{theorem} \label{thm:main-lb} Consider the set of functions $\bF_C$ (defined over a fixed mesh $\MM$) 
% that all have the same value $C$ as a valid cost.  Any height function decision tree that correctly computes
% the contour tree on all of $\bF_C$ has an average case (uniform over $\bF_C$) running time of $\Omega(C)$.
% \end{theorem}

In other words, it is not possible to improve on \Thm{main-alg} even when restricted to a subset
of inputs of fixed cost.

@@

There has been a long series of improved algorithms for computing contour trees.
van Kreveld \etal \cite{kobps-ctsssit-97} presented an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N)$.
One cannot overstate the importance of this result, since it forms the basis 
of subsequent practical and theoretical work in computational topology applications. 
Cole-McLaughlin and Pascucci \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} improve upon this result, and provide an unconditional $O(N+t\log t)$ algorithm.

@@

\begin{figure}[h]\centering
    \includegraphics[width=.7\linewidth,height=.3\linewidth]{figs/balanced4}
    \vspace{.1in}

@@

This is convenient, as we can now assume that $f$ is defined on $\RR^d$. Any point inside a boundary facet
has a well-defined height, and this includes the infinite facet. 
We think of $d$ as constant, and assume that $\MM$ is represented in a data structure that allow constant-time traversals of neighboring simplices
in $\MM$ (e.g.~\cite{BoMa12}). (This is analogous to a DCEL for higher dimensions.)
Observe that $f:\MM \rightarrow \RR$ can be thought of as a $d$-dimensional simplicial complex living in $\RR^{d+1}$.
We can think of $f(x)$ as the ``height" of a point $x \in \MM$, and encoded in the representation of $\MM$. 
%Abusing notation, $\MM$ refers to this encoding, so we do not refer to $f$.

@@

\begin{definition} \label{def:tree} The contour tree is the graph on vertex set $\cV$, where
edges are formed as follows. For every contour class that is created at $v_i$ and destroyed $v_j$,
there is an edge $(v_i,v_j)$. (Conventionally, edges are directed from higher to lower function value.)
\end{definition}

We denote the contour tree of $\MM$
by $\reeb(\MM)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
It is not immediately obvious that this graph is a tree, but alternate definitions of the contour tree
in~\cite{csa-cctad-00} imply this is a tree. 
Since this tree has height values associated with the vertices, we can talk about up-degrees and down-degrees in $\reeb(\MM)$.
We assume there are no ``multi-saddles", so up and down-degrees are at most $2$, and the total degree is at most $3$. 
This is again a standard assumption in topological algorithms and can be achieved by vertex unfolding (Section 6.3 in~\cite{HaEd10}).

@@

The $(f(x)+\delta)$-contours in $f|_{B_\eps(x)}$ given by \Def{deg} might actually be the same
contour in $f$. The up-degree is defined as the number of $(f(x)+\delta)$-contours
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, then the level set gains a handle.
When $d=2$, this distinction is not necessary, since any point with $>1$
Morse degree will have some $>1$ degree in $\reeb(\MM)$.

@@

\smallskip
\begin{asparaenum}
	\item Let $\MM' = \cut(\MM,\phi)$.
	\item Construct $\reeb(\MM')$ and let $A, B$ be the nodes corresponding to the new boundaries created
	in $\MM'$. (One is a minima and the other is maxima.)
	\item Since $A, B$ are leaves, they each have unique neighbors $A'$ and $B'$, respectively. Insert
	edge $(A',B')$ and delete $A, B$ to obtain $\reeb(\MM)$.
\end{asparaenum}
\end{minipage}}

@@

\begin{lemma} \label{lem:cut} Consider a regular contour $\phi$ contained in a contour class (of an edge of $\reeb(\MM))$
$(u,v)$ and denote $\MM' = \cut(\MM,\phi)$. Then $\cV(\reeb(\MM')) = \{\phi^+,\phi^-\} \cup \cV(\MM)$
and $\cE(\reeb(\MM')) = \{(u,\phi^+),(\phi^-,v)\} \cup (\cE(\MM) \setminus (u,v))$.
\end{lemma}

@@

\begin{theorem} \label{thm:jordan} $\cut(\MM,\phi)$ consists of two
disconnected simplicial complexes.
\end{theorem}

\begin{proof} Denote (as in \Lem{cut}) the edge containing $\phi$ to be $(u,v)$. Suppose for contradiction that there is a path between vertices $u$ and $v$
in $\MM' = \cut(\MM,\phi)$. By \Thm{carr-path}, there is a path in $\cC(\MM')$ between $u$ and $v$. Since $\phi^+$ and $\phi^-$
are leaves in $\cC(\MM')$, this path obviously cannot use edges incident to them. By \Lem{cut},
all the edges of this path are in $\cE(\cC(\MM)) \setminus (u,v)$. So we get a cycle in $\cC(\MM)$, a contradiction.
To show that there are exactly two connected components in $\cut(\MM,\phi)$, it suffices
to see that $\cC(\MM')$ has two connected components (by \Lem{cut}) and apply \Thm{carr-path}.
\end{proof}

@@

\begin{figure}[h]\centering
    \includegraphics[width=.25\linewidth]{figs/tent2}%
    \hspace{.9in}
    \includegraphics[width=.25\linewidth]{figs/tent1}
    \caption{On left, downward paint spilling only (each shade of gray represents a different color), producing a grid.  On right, flipping the direction of paint spilling.}
    \label{fig:order}
\end{figure}

@@

\smallskip
\begin{compactenum}
	\item Determine interface of $\wet(x,\NN)$. 
	\item If the interface is empty, simply output $\NN$. Otherwise, denote the contours by $\phi_1, \phi_2, \ldots, \phi_k$ and set $\phi'_i = \lift(\phi_i)$.
	\item Initialize $\NN_1 = \NN$.
	\item For $i$ from $1$ to $k$: 
	\begin{compactenum}
		\item Construct $\cut(\NN_i,\phi'_i)$, consisting of dry complex $\LL_i$ and remainder $\NN_{i+1}$.
		\item Let the newly created boundary of $\LL_i$ be $B_i$. Invert $\LL_i$ so that $B_i$ is a maximum. Recursively
		call $\rain(B_i,\LL_i)$.
	\end{compactenum}
	\item Output $\NN_{k+1}$ together with any complexes output by recursive calls.
\end{compactenum}
\end{minipage}}

@@

Consider the first time that $F$ is cut, say, during the call to $\rain(x,\NN)$.
Specifically, say this happens when $\cut(\NN_i,\phi'_i)$ is constructed. 
So $F$ is cut by some hyperplane into lower and upper portions (which are then triangulated).
Since the cut is infinitesimally above an interface, the lower portion is basically wet.
It can never be part of $\LL_1, \LL_2, \ldots$, which go into recursive calls. This portion
is in $\NN_{k+1}$. The upper portion/facet (call it $U$) is in $\LL_i$. Note that the lower
boundary of $U$ is in the boundary $B_i$. Since a recursive call is made to $\rain(B_i,\LL_i)$
(and $\LL_i$ is inverted), $U$ becomes wet. Hence, $U$ will not be cut subsequently.
All in all, $F$ is cut at most once.
\ignore{
As the number of edges is $O(n)$, arguing that there is at most one boundary vertex created along any given edge will imply the claim.
Consider some edge $e$ and take the first time a boundary vertex is created in $e$.
This happens when some contour $\phi$ intersects $e$. On applying $\cut(\NN,\lift(\phi))$,
consider the dry manifold $\LL$ and wet manifold $\MM$ constructed. The edge $e$ is split into two parts, the dry and wet part.
The wet part is in $\MM$ and will never be involved in any recursive call.

@@

\section{Painting to compute contour trees}
\label{sec:paint}

@@

\begin{definition}
\label{def:leafAssign}
 A \emph{leaf assignment} $\chi$ of a tree $T$ assigns \emph{two} distinct leaves to each internal vertex $v$,
 one from the left child and one from the right child subtree of $v$ (naturally if $v$ has only one child it is assigned 
 only one color).
\end{definition}

@@

\newcommand{\pathTree}{P_{\mathcal{S}}}
\newcommand{\pathTreeA}{\mathcal{P}_{\mathcal{S}}}
\section{Leaf assignments and path decompositions}
\label{sec:pathDecomp}

@@

\subsection{Join Trees}

In this section we prove \Thm{main-lb}, but for join trees instead of contour trees.  Observe that given a fixed $(f,\MM)$, 
our algorithm for computing $\jc(\MM)$ is deterministic and so the path decomposition $P=P(\jc(\MM))$ in \Thm{runtime} is fixed, 
allowing us to define the cost, $\cost(f)=\sum_{p\in P} |p|\log|p|$.

%In the following note that two distinct terrains can have the same graph structure defining their meshes, but different height values associated with their vertices.

@@

\begin{lemma}
\label{lem:pathDecomp}
Let $\cost(f)=\sum_{p\in P} |p|\log|p|$, for some height function $f$ defined over a complex $\MM$.
Then there is a family of functions, $F$, all defined over a common triangulation in plane, 
such that the size of $F$ is $\Pi_{p_i\in P} (|p_i|-1)!$, and no two function in $F$ define the same join tree.
\end{lemma}
\begin{proof}
 We describe a simple triangulation, with heights specified for all vertices except for the saddles.  The main idea is that 
 this triangulation will be flexible enough to allow almost any assignment of heights to the saddles, thus defining a class of join trees 
 whose size is that described in the lemma statement.
